
## Step Trends by Day of Week and Time

#### 0) Load the data into R

```{r, cache=TRUE}
setwd('C:\\Users\\Nic\\Desktop\\Nic\\Uni Courses\\coursera\\DS Specialisation\\reproducible research\\assign')
dat <- read.csv('activity.csv', header=TRUE)

head(dat)
```

#### 1) Mean total number of steps taken per day

The histogram of the total number of steps taken per day ignores cases of missing steps. These values are imputed in a section below.


```{r}
step_day_summary <- tapply(dat$steps, dat$date, sum, na.rm = TRUE)

hist(step_day_summary, breaks = 50, col = "pink",
     xlab = "Mean Total Number of Steps per Day")
```

```{r}
mean_total_daily_steps = as.integer(mean(step_day_summary, na.rm = TRUE)) 
median_total_daily_steps = as.integer(median(step_day_summary, na.rm = TRUE)) 
```

The mean and median number of steps taken per day are:  
* mean: `r mean_total_daily_steps`  
* median: `r median_total_daily_steps`  

Which interestingly shows many days where very few steps were taken. This is questionable and may be related to missing valuesfor that particular day.

#### 2) What is the average daily activity pattern?

```{r}
step_time_summary <- tapply(dat$steps, dat$interval, mean, na.rm=TRUE)

y = as.integer(row.names(step_time_summary)) # this indicates the time in 5 minute intervals

plot(y, step_time_summary, type="l",
     main = "Average Steps by 5 min Time Intervals
     Across All Days", xlab = "Time (5 min intervals)")

max_steps = data.frame(step_time_summary[step_time_summary == max(step_time_summary)])

```

The maximum average steps in a 5 minute time interval occurs at `r labels(max_steps)[1]` and is `r max_steps` steps.

#### 3) Imputing missing values

```{r}
# total number of missing values calc
miss = sum(is.na(dat$steps))
```

The total number of missing values in the dataset is `r miss`.

These missing values can be imputed by taking the mean of the non-missing 5 minute intervals in place of the missing data.

```{r}
t <- data.frame(table(dat$interval[is.na(dat$steps)==TRUE]))
head(t) # examine some of the missing data
```

As it turns out we have 8 missing values for each time interval.

Here we are calculating a table of times and mean steps for each time interval.

```{r}
impute_table <- data.frame(tapply(dat$steps, dat$interval, mean, na.rm = TRUE))
impute_table$interval <- row.names(impute_table)
names(impute_table)[1] = "steps_non_missing"
row.names(impute_table) <- NULL
hist(impute_table$steps_non_missing, breaks = 50,
     main = "Non-missing mean steps for each time interval", xlab = "Steps per Interval")
impute_table$interval <- as.integer(impute_table$interval)
```

Here is the dataset with the missing data imputed

```{r, cache=FALSE}
library(plyr)
```

```{r}
dat <- join(dat, impute_table, type = "left")
dat$steps <- ifelse(is.na(dat$steps),dat$steps_non_missing, dat$steps)
hist(dat$steps, breaks = 50, main = "Mean steps with missing values imputed", xlab = "Mean Steps per Interval")
```

Histogram of total number of steps taken each day with missing values imputed 

```{r}
step_day_summary <- tapply(dat$steps, dat$date, sum)

hist(step_day_summary, breaks = 50, main = "Mean Total Number of steps taken per day", xlab = "Mean Total Steps per Day")

mean_imputed = as.integer(mean(step_day_summary))
median_imputed = as.integer(median(step_day_summary))
```

When we impute missing values we have the mean number of total steps taken per day is `r mean_imputed` and the median number of steps taken per day is `r median_imputed`  

#### 4) Differences in Activity Pattern Between Weekdays and Weekend

```{r, cache = TRUE}
library(lubridate)
library(lattice)
```

We have flagged weekend days and weekdays below.

```{r}
datey <- ymd(dat$date)
dat$weekday_flag = ifelse(weekdays(datey) %in% c('Saturday','Sunday'), 'Weekend', 'Weekday')
table(dat$weekday_flag)
```

So now we can see the 8 days with very few steps have had steps imputed. These days now fall towards the average of the distribution (as we would expect).

```{r}
datmean <- ddply(dat[,c('weekday_flag', 'interval', 'steps')], .(weekday_flag, interval), summarise,  ms = mean(steps))

xyplot(ms ~ interval | weekday_flag, data = datmean, type = "l", main = "Mean steps per day cut by Weekday and Weekend", xlab = "Time", ylab = "Mean Steps")
```

Of a weekday we see trends in activity as people presumably get up and go to work, of a late afternoon and evening we see activity as people return home. During the day people look more sedentary.

Of a weekend people may rise at later times, so we miss the spike in morning activity. Although during the day people appear to be less sedentary.